Overview

### 2023-12-02 analysis/001_overview.Rmd
Journal investigated: hybrid journals with at least one oa article between 2018 - 2022 (check 0001-4338)

---
title: "Overview"
output: github_document
date: "`r Sys.Date()`"
---

@@

```{r dependencies}
library(tidyverse, warn.conflicts = FALSE)
```

@@

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  #   filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

@@

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023"), !is.na(cc)) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

@@

Between 2018 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` out of `r n_total` articles, representing a five-year hybrid open access uptake rate of `r n_oa / n_total * 100`%.
First authors affiliated with institutions that had a transformative agreement in place at the time of publication, which indicates that they could make use of such agreements, contributed `r n_ta_oa` open access articles to the total.

@@

```{r journal_boxplot}
jn_hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi)) |>
  mutate(cr_year = factor(cr_year))
# year all
jn_year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424") |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year, issn_l) |> 
  summarise(n = sum(jn_all))

@@

Figure 2A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall uptake rate and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, the overall open access uptake rate increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
At the same time, the total publication volume of the investigated journals grew from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure 2B highlights that the majority of open access articles in hybrid journals were made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access volume in 2022.
However, there was also a notable growth in open access provision through individual publication fees, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). This suggests that publishers were able to gain equally from individual and institutional open access publishing options.

@@

jn_ind_jct_all <- jn_ind_jct |>
  distinct(cr_year, issn_l, esac_publisher, jn_all)

@@

## plot
publisher_plot_df <- publisher_df |> 
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  mutate(
    publisher = case_when(
      esac_publisher == "Elsevier" ~ "Elsevier",
      esac_publisher == "Springer Nature" ~ "Springer Nature",
      esac_publisher == "Wiley" ~ "Wiley",
      is.character(esac_publisher) ~ "Other"
    )
  ) |>
  mutate(publisher =
           forcats::fct_relevel(
             publisher,
             c("Elsevier", "Springer Nature", "Wiley", "Other") 
           )) |>
  filter(!cr_year %in%  c("2017", "2023"), oa > 0)

@@

publisher_boxplot <- publisher_plot_df |>
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value / jn_all, fill = type)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement")) +
  coord_cartesian(ylim = c(0,0.65)) +
  facet_wrap(~publisher, nrow = 1) +
  labs(y = "OA Uptake\nby Hybrid Journal", x = "Publication Year") +
  theme_minimal() +
    theme(strip.background = element_blank(),
    strip.text.x = element_blank()
    )

@@

```{r, fig.asp=1.1}
library(patchwork)
design = "A
B
C"
wrap_plots(
  A = publisher_prop,
  B = publisher_ta_prop_plot,
  C = publisher_boxplot,
  design = design,
  guides = "collect",
 # widths = c(2, 1),
  heights = c(1, 0.5, 1)
) &
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10)) &
    plot_annotation(tag_levels = 'A') 
```

>>>>

---
title: "Overview"
output: github_document
always_allow_html: true
date: "`r Sys.Date()`"
---

@@

```{r dependencies}
library(tidyverse, warn.conflicts = FALSE)
library(kableExtra)
```

@@

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

@@

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

@@

Between 2018 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` out of `r n_total` articles, representing a five-year hybrid open access uptake of `r n_oa / n_total * 100`%.
First authors affiliated with institutions that had a transformative agreement in place at the time of publication, which indicates that they could make use of such agreements, contributed `r n_ta_oa` open access articles to the total.

@@

```{r journal_boxplot}
jn_hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424") |>
  group_by(cr_year, issn_l) |>
  summarise(oa_all = n_distinct(doi)) |>
  mutate(cr_year = factor(cr_year))
# year all
jn_year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424") |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year, issn_l) |> 
  summarise(n = sum(jn_all))

@@

Figure 2A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
At the same time, the total article volume of the investigated journals grew from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure 2B highlights that the majority of open access articles in hybrid journals were made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable growth in open access provision through individual publication fees, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). This suggests that publishers were able to gain equally from individual and institutional open access publishing options.

@@

jn_ind_jct_all <- jn_ind_jct |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(cr_year, issn_l, esac_publisher, jn_all)

@@

## plot
publisher_plot_df <- publisher_df |> 
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  mutate(
    publisher = case_when(
      esac_publisher == "Elsevier" ~ "Elsevier",
      esac_publisher == "Springer Nature" ~ "Springer Nature",
      esac_publisher == "Wiley" ~ "Wiley",
      is.character(esac_publisher) ~ "Other"
    )
  ) |>
  mutate(publisher =
           forcats::fct_relevel(
             publisher,
             c("Elsevier", "Springer Nature", "Wiley", "Other") 
           )) |>
  filter(!cr_year %in%  c("2017", "2023"))

@@

publisher_boxplot <- publisher_plot_df |>
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value / jn_all, fill = type)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE, varwidth = TRUE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement")) +
  coord_cartesian(ylim = c(0,0.65)) +
  facet_wrap(~publisher, nrow = 1) +
  labs(y = "OA Uptake\nby Hybrid Journal", x = "Publication Year") +
  theme_minimal() +
    theme(strip.background = element_blank(),
    strip.text.x = element_blank()
    )

@@

Analysing hybrid open access across publishers between 2018 and 2022 reveals a large concentration on a few publishers in terms of publication volume. 
While transformative agreements could be attributed to `r length(unique(publisher_plot_df$esac_publisher))` distinct publishers, the three publishers Elsevier, Springer Nature and Wiley accounted for xxx percent of articles published. Together, they published xxx articles, representing a share of xx percent. xxx were published through transformative agreements. 
The remainder consists of

```{r publisher_summary_table}
publisher_article_total_table <- publisher_summary_df |>
  distinct(publisher, jn_all) |>
  group_by(publisher) |>
  summarise(publisher_total = sum(jn_all)) |>
  mutate(perc = publisher_total / sum(publisher_total),
         cum = cumsum(publisher_total) / sum(publisher_total))

publisher_article_oa_table <- publisher_summary_df |>
  group_by(publisher) |>
  summarise(oa_total = sum(value)) |>
  mutate(oa_perc = oa_total / sum(oa_total),
         oa_cum = cumsum(oa_total) / sum(oa_total))

publisher_article_ta_oa_table <- publisher_summary_df |>
  filter(type == "ta_oa") |>
  group_by(publisher) |>
  summarise(ta_oa_total = sum(value)) |>
  mutate(ta_oa_perc = ta_oa_total / sum(ta_oa_total),
         ta_oa_cum = cumsum(ta_oa_total) / sum(ta_oa_total))

publisher_league_tbl <-
  inner_join(publisher_article_total_table, publisher_article_oa_table) |>
  inner_join(publisher_article_ta_oa_table)

kbl(publisher_league_tbl) %>%
  kable_classic() %>%
  add_header_above(c("Publisher" = 1, "Group 1" = 3, "Group 2" = 3, "Group 3" = 3))
```

Figure 2 presents the growth of hybrid open access across publishers by year, highlighting the proportion of open articles published under transformative agreements. As can be seen from Figure 2A and 2C

```{r, fig.asp=1.1}
library(patchwork)
design = "E
A
B
C"
wrap_plots(
  E = guide_area(),
  A = publisher_prop,
  B = publisher_ta_prop_plot,
  C = publisher_boxplot,
  design = design,
  guides = "collect",
  heights = c(0, 0.8, 0.3, 0.8)
) &
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10)) &
    plot_annotation(tag_levels = 'A') 
```
